Exploring the composition of lithic assemblages in Mesolithic south-eastern Norway

### 2021-07-23 analysis/paper/paper.Rmd
Working through David's comments

The lithic data analysed here is based on the classification of the site assemblages done for the original excavation reports, and consists of `r ncol(select(sites, core:flake_retouch_other))` variables representing different debitage and tool types. While the classification practices for the excavation projects were standardised to an extent, there are some instances where time was allocated to identify additional artefact sub-categories aimed at answering specific research questions. Some categories in the original reports have therefore been combined in the dataset. This for example pertains to the category narrow-blades (width \> 8mm and \< 12mm), which was only separated from (macro-)blades (width $\ge$ 12 mm) and micro-blades (width $\le$ 8 mm) for some of the sites. Narrow-blades were combined with the blade category here. Furthermore, the artefact data have here been divided into flint and non-flint materials. Flint does not occur indigenously in Norway, and is only available locally as nodules that have been transported and deposited by retreating and drifting ice [e.g. @berg-hansen1999]. This means that the distribution and quality of flint has been impacted by a diverse set of factors relating to climatic and geographical factors such as, but not limited to, topographic variability, shoreline morphology and ocean currents [@eigeland2015, 46]. Thus, while flint is treated as a unified category here, the variability in quality could have been substantial [@eigeland2015, 45-53]. Furthermore, the various non-flint raw materials that have been lumped together have quite disparate properties, where fine-grained cryptocrystalline materials are often used as a substitute or supplement to flint, while other, coarser materials are usually associated with the production of axes and other macro tools. Given this differentiated use, these raw-material properties are expected to be reflected in the retained debitage and tool categories. An important benefit of combining all of the non-flint materials is that this reduces the dependency on whether or not these have been correctly and consistently categorised for the reports [cf. @frivoll2017]. While certainly a topic deserving of more attention, the general sentiment in the literature is that there would have been stable access to locally available non-flint raw-materials of good quality in south-eastern Norway [e.g. @eigeland2015, 370; @glørstad2011]. Finally, while factors such as landscape changes through shoreline displacement can have led to variable raw-material availability at the analysed sites, their relatively constrained geographical distribution hopefully counteracts some non-behavioural sources of variation.

Studies concerned with chronological changes in the composition of lithic assemblages in southern Norway have typically had a focus on morphological variation among artefacts [e.g. @ballin1999; @bjerck1986; @reitan2016] or been concerned with technological processes associated with certain sub-categories of the site inventories, such as the production of blades or axes [e.g. @berg-hansen2017; @eymundsson2018; @damlien2016; @solheim2020]. Studies that have involved entire assemblages have either been concerned with general compositional traits such as relative frequency of various tool types and raw-materials [@breivik2016; @viken2018c; @breivik2020; @reitan2016], or involved extremely in-depth studies of technological organisation associated with a handful of assemblages [@fuglestvedt2007; @eigeland2015; @mansrud2016]. These studies are, however, based on narratively driven methods, leaving the weighting of the different variables for the final interpretations unclear. To my knowledege, only a single study dealing with the composition of Mesolithic assemblages in southern Norway has involved the use of a multivariate quantitative framework, which was employed to structure the analysis of eight Middle Mesolithic assemblages [@solheim2013a; see @glørstad2010, 145--146 for a spatial application]. In sum then, previous studies have typically either been limited to a small number of sites, to a subset of the inventories, to morphological characteristics, or to subjectively and narratively driven methods that are difficult to scale and consistently balance in the comparison of a larger number of artefact categories and assemblages.

@@

To get more directly at this temporal trend, a curation index based on VDL and RFSL was devised by first performing a min-max normalisation of the two variables, scaling them to take on values between 0 and 1. The values for artefact density was then made negative to reflect its relationship with degree of curation. The mean was then found for each site on these two normalised values. To account for the temporal uncertainty associated with the dating of the sites, a simulation-based approach was also adopted [e.g. @baxter2016; @crema2012; @orton2017]. A LOESS curve was fit to the curation index and site age for each simulation run, where the age of each site was drawn as a single year from the date ranges associated with the sites as provided in Figure \@ref(fig:dist). For sites with radiocarbon age determinations the dates were drawn from the associated summed posterior density estimates, while ages for sites dated with reference to relative sea-level change and typology were drawn uniformly from the associated date range. This simulation was repeated 1000 times, the results of which is visualised in Figure \@ref(fig:simplot)A. Disregarding the edge-effects at either end of the plot, the general tendency is a relatively high degree of curation among the earlier sites, followed by a marked drop around 8000 BCE. This has stabilised by around 7000 BCE and remains stable without any major fluctuations for the rest of the Mesolithic. The variation in degree of curation is also markedly higher after 8000 BCE, potentially reflecting variation in associated mobility patterns. Figure \@ref(fig:simplot)B displays the result of running the same procedure on the flint data. The general pattern follows the same trajectory, but the result for some individual sites is markedly different. This is discussed below.

@@

The relevance of the relationship between frequency of secondarily worked lithics and volumetric density of lithics was here identified by means of an exploratory approach, and is in part investigated further because they align with suggestions from previous research --- clearly representing a possible case of both confirmation bias and circular reasoning. However, some inferential merit can be achieved by invoking what has been termed consilience. Consilience involves "the interlocking or coherence of causal explanations across multiple problem domains" through a clear operationalisation of explanatory terms and concepts [@clark2009, 30]. Thus, the overlap in results presented here and those repeatedly reported from a range of different context does speak to the general applicability and comparability offered by the measures, and gives an initial indication that they might be capturing the social dimensions of interest also in a Norwegian Mesolithic setting.

>>>>

The lithic data analysed here is based on the classification of the site assemblages done for the original excavation reports, and consists of `r ncol(select(sites, core:flake_retouch_other))` variables representing different debitage and tool types. While the classification practices for the excavation projects were standardised to an extent, there are some instances where time was allocated to identify additional artefact sub-categories aimed at answering specific research questions. Some categories in the original reports have therefore been combined in the dataset. For example, this pertains to the category narrow-blades (width \> 8mm and \< 12mm), which was only separated from (macro-)blades (width $\ge$ 12 mm) and micro-blades (width $\le$ 8 mm) for some of the sites. Narrow-blades were combined with the blade category here. Furthermore, the artefact data have here been divided into flint and non-flint materials. Flint does not occur indigenously in Norway, and is only available locally as nodules that have been transported and deposited by retreating and drifting ice [e.g. @berg-hansen1999]. This means that the distribution and quality of flint has been impacted by a diverse set of factors relating to climatic and geographical factors such as, but not limited to, topographic variability, shoreline morphology and ocean currents [@eigeland2015, 46]. Thus, while flint is treated as a unified category here, the variability in quality could have been substantial [@eigeland2015, 45-53]. Furthermore, the various non-flint raw materials that have been lumped together have quite disparate properties, where fine-grained cryptocrystalline materials are often used as a substitute or supplement to flint, while other, coarser materials are usually associated with the production of axes and other macro tools. Given this differentiated use, these raw-material properties are expected to be reflected in the retained debitage and tool categories. An important benefit of combining all of the non-flint materials is that this reduces the dependency on whether or not these have been correctly and consistently categorised for the reports [cf. @frivoll2017]. While certainly a topic deserving of more attention, the general sentiment in the literature is that there would have been stable access to locally available non-flint raw-materials of good quality in south-eastern Norway [e.g. @eigeland2015, 370; @glørstad2011]. Finally, while factors such as landscape changes through shoreline displacement can have led to variable raw-material availability at the analysed sites, their relatively constrained geographical distribution hopefully counteracts some non-behavioural sources of variation.

Studies concerned with chronological changes in the composition of lithic assemblages in southern Norway have typically had a focus on morphological variation among artefacts [e.g. @ballin1999; @bjerck1986; @reitan2016] or been concerned with technological processes associated with certain sub-categories of the site inventories, such as the production of blades or axes [e.g. @berg-hansen2017; @eymundsson2018; @damlien2016; @solheim2020]. Studies that have involved entire assemblages have either been concerned with general compositional traits such as relative frequency of various tool types and raw-materials [@breivik2016; @viken2018c; @breivik2020; @reitan2016], or involved extremely in-depth studies of technological organisation associated with a handful of assemblages [@fuglestvedt2007; @eigeland2015; @mansrud2016]. These studies are, however, based on narratively driven methods, leaving the weighting of the different variables for the final interpretations unclear. To my knowledge, only a single study dealing with the composition of Mesolithic assemblages in southern Norway has involved the use of a multivariate quantitative framework, which was employed to structure the analysis of eight Middle Mesolithic assemblages [@solheim2013a; see @glørstad2010, 145--146 for a spatial application]. In sum then, previous studies have typically either been limited to a small number of sites, to a subset of the inventories, to morphological characteristics, or to subjectively and narratively driven methods that are difficult to scale and consistently balance in the comparison of a larger number of artefact categories and assemblages.

@@

To get more directly at this temporal trend, a curation index based on VDL and RFSL was devised by first performing a min-max normalisation of the two variables, scaling them to take on values between 0 and 1. The values for artefact density was then made negative to reflect its relationship with degree of curation. The mean was then found for each site on these two normalised values. To account for the temporal uncertainty associated with the dating of the sites, a simulation-based approach was also adopted [e.g. @baxter2016; @crema2012; @orton2017]. A LOESS curve was fit to the curation index and site age for each simulation run, where the age of each site was drawn as a single year from the date ranges associated with the sites as provided in Figure \@ref(fig:dist). For sites with radiocarbon age determinations the dates were drawn from the associated summed posterior density estimates, while ages for sites dated with reference to relative sea-level change and typology were drawn uniformly from the associated date range. This simulation was repeated 1000 times, the results of which is visualised in Figure \@ref(fig:simplot)A. Disregarding the edge-effects at either end of the plot, the general tendency is a relatively high degree of curation among the earlier sites, followed by a marked drop around 8000 BCE. This has stabilised by around 7000 BCE and remains stable without any major fluctuations for the rest of the Mesolithic. The variation in degree of curation is also markedly higher after 8000 BCE, potentially reflecting variation in associated mobility patterns. Figure \@ref(fig:simplot)B displays the result of running the same procedure on the flint data. The general pattern follows the same trajectory, but the result for some individual sites is markedly different.

@@
